A comparison of design-based and model-based approaches for finite population spatial sampling and inference.

### 2022-05-19 inst/manuscript/manuscript.tex
knitted version with correct citations.

\usepackage{lineno} % add
  \linenumbers % turns line numbering on

\usepackage{graphicx}
%%%%%%%%%%%%%%%% end my additions to header

@@

% tightlist command for lists without linebreak
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

@@

\begin{document}

\begin{frontmatter}

@@

Spatial data are data that have some sort of spatial index (usually
specified via coordinates). De Gruijter and Ter Braak (1990) and Brus
and DeGruijter (1993) give early comparisons of design-based and
model-based approaches for spatial data, quashing the belief that
design-based approaches could not be used for spatially correlated data.
Since then, there have been several general comparisons between
design-based and model-based approaches for spatial data (Brus and De
Gruijter, 1997; Brus, 2021; Ver Hoef, 2002, 2008). Cooper (2006) reviews
the two approaches in an ecological context before introducing a
``model-assisted'' variance estimator that combines aspects from each
approach. In addition to Cooper (2006), there has been substantial
research and development into estimators that use both design-based and
model-based principles (see e.g., Sterba (2009) and Cicchitelli and
Montanari (2012), and for Bayesian approaches, see Chan-Golston et al.
(2020) and Hofman and Brus (2021)).

@@

When samples are chosen in a manner such that the layout of sampled
units reflects the layout of the population units, we call the resulting
sample spatially balanced. By ``reflecting the layout of the population
units'', we mean that if population units are concentrated in specific
areas, the units in the sample should be concentrated in the same areas.
Because spatially balanced samples reflect the layout of the population
units, they are not necessarily spread out in space in some equidistant
manner. One method of selecting spatially balanced samples is the
generalized random tessellation stratified (GRTS) algorithm (Stevens and
Olsen, 2004), which we discuss in more detail in Section
\ref{subsec:spb_design}. To quantify the spatial balance of a sample,
Stevens and Olsen (2004) proposed loss metrics based on Voronoi polygons
(i.e., Dirichlet Tessellations).

@@

Finite population block kriging (FPBK) is a model-based approach that
expands the geostatistical Kriging framework to the finite population
setting (Ver Hoef, 2008). Instead of developing inference based on a
specific sampling design, we assume the data are generated by a spatial
stochastic process. We summarize some of the basic principles of FPBK
next -- see Ver Hoef (2008) for technical details and see Higham, Ver
Hoef, Madsen, et al. (2021) for an extension to cases of imperfect
detection among population units. Let
\({\mathbf{z} \equiv \{\text{z}(s_1), \text{z}(s_2), . . . , \text{z}(s_N) \}}\)
be an \(N \times 1\) response vector at locations \(s_1\), \(s_2\), . .
. , \(s_N\) that can be measured at the \(N\) population units. Suppose
we want to use a sample to predict some linear function of the response
variable, \(f(\mathbf{z}) = \mathbf{b}^\prime \mathbf{z}\), where
\(\mathbf{b}^\prime\) is a \(1 \times N\) vector of weights (e.g, the
population mean is represented by a weights vector whose elements all
equal \(1 / N\)). Denoting quantities that are part of the sampled
population units with a subscript \emph{s} and quantities that are part
of the unsampled population units with a subscript \emph{u}, let

@@

In this section we describe how we used simulated and real data to
investigate performance between simple random sampling (SRS) and GRTS
sampling as well as performance between design-based (DB) and
model-based (MB) inference. In SRS and GRTS sampling, all population
units had equal inclusion probabilities and were selected without
replacement. The important distinction between SRS and GRTS is that SRS
ignores spatial locations while sampling but GRTS explicitly
incorporates them. Together, the two sampling plans (SRS and GRTS)
combined with the two inference approaches (DB and MB) yielded four
sampling-inference combinations: SRS-DB, SRS-MB, GRTS-DB, and GRTS-MB.
For SRS-DB, the Horvitz-Thompson estimator \eqref{eq:ht} was used to
estimate means and the commonly-used SRS variance formula (Lohr, 2009;
Särndal et al., 2003) was used to estimate variances. This variance
formula is given by \begin{equation}\label{equation:srs_var}
 \frac{f[\sum_{i = 1}^n (z_i - \bar{z})^2]}{n(n - 1)},
\end{equation} where \(z_i\) is the response value of the \(i\)th
sampled population unit, \(\bar{z}\) is the mean of all \(z_i\), \(n\)
is the sample size, \(N\) is the population size, and
\(f = (1 - n / N)\) (\(f\) is often called the finite population
correction factor). For GRTS-DB, the Horvitz-Thompson estimator was used
to estimate means and the local neighborhood variance was used to
estimate variances. For SRS-MB and GRTS-MB, FPBK was used to estimate
means and variances using restricted maximum likelihood. SRS, GRTS
sampling, and design-based inference were implemented using the
\texttt{spsurvey} \textbf{\textsf{R}} package (Dumelle et al., 2022).
FPBK was implemented using the \texttt{sptotal} \textbf{\textsf{R}}
package (Higham, Ver Hoef, Frank, et al., 2021).

@@

\leavevmode\hypertarget{ref-higham2021sptotal}{}%
Higham, M., Ver Hoef, J., Frank, B., Dumelle, M., 2021. Sptotal:
Predicting totals and weighted sums from spatial data.

\leavevmode\hypertarget{ref-higham2021adjusting}{}%
Higham, M., Ver Hoef, J., Madsen, L., Aderman, A., 2021. Adjusting a
finite population block kriging estimator for imperfect detection.
Environmetrics 32, e2654.

>>>>

\usepackage{lineno} % add
  \linenumbers % turns line numbering on
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\usepackage{graphicx}
\usepackage{booktabs} % book-quality tables
%%%%%%%%%%%%%%%% end my additions to header

@@

% Pandoc citation processing
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
% for Pandoc 2.8 to 2.10.1
\newenvironment{cslreferences}%
  {}%
  {\par}
% For Pandoc 2.11+
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
  % set entry spacing
  \ifnum #2 > 0
  \setlength{\parskip}{#2\baselineskip}
  \fi
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

% Pandoc header

@@

\begin{document}
\begin{frontmatter}

@@

Spatial data are data that have some sort of spatial index (usually
specified via coordinates). De Gruijter and Ter Braak (1990) and Brus
and DeGruijter (1993) give early comparisons of design-based and
model-based approaches for spatial data, quashing the belief that
design-based approaches could not be used for spatially correlated data.
Since then, there have been several general comparisons between
design-based and model-based approaches for spatial data (Brus and De
Gruijter, 1997; Brus, 2021; Ver Hoef, 2002; Ver Hoef, 2008). Cooper
(2006) reviews the two approaches in an ecological context before
introducing a ``model-assisted'' variance estimator that combines
aspects from each approach. In addition to Cooper (2006), there has been
substantial research and development into estimators that use both
design-based and model-based principles (see e.g., Sterba (2009) and
Cicchitelli and Montanari (2012), and for Bayesian approaches, see
Chan-Golston et al. (2020) and Hofman and Brus (2021)).

@@

When samples are chosen in a manner such that the layout of sampled
units reflects the layout of the population units, we call the resulting
sample spatially balanced. By ``reflecting the layout of the population
units,'' we mean that if population units are concentrated in specific
areas, the units in the sample should be concentrated in the same areas.
Because spatially balanced samples reflect the layout of the population
units, they are not necessarily spread out in space in some equidistant
manner. One method of selecting spatially balanced samples is the
generalized random tessellation stratified (GRTS) algorithm (Stevens and
Olsen, 2004), which we discuss in more detail in Section
\ref{subsec:spb_design}. To quantify the spatial balance of a sample,
Stevens and Olsen (2004) proposed loss metrics based on Voronoi polygons
(i.e., Dirichlet Tessellations).

@@

Finite population block kriging (FPBK) is a model-based approach that
expands the geostatistical Kriging framework to the finite population
setting (Ver Hoef, 2008). Instead of developing inference based on a
specific sampling design, we assume the data are generated by a spatial
stochastic process. We summarize some of the basic principles of FPBK
next -- see Ver Hoef (2008) for technical details and see Higham et al.
(2021b) for an extension to cases of imperfect detection among
population units. Let
\({\mathbf{z} \equiv \{\text{z}(s_1), \text{z}(s_2), . . . , \text{z}(s_N) \}}\)
be an \(N \times 1\) response vector at locations \(s_1\), \(s_2\), . .
. , \(s_N\) that can be measured at the \(N\) population units. Suppose
we want to use a sample to predict some linear function of the response
variable, \(f(\mathbf{z}) = \mathbf{b}^\prime \mathbf{z}\), where
\(\mathbf{b}^\prime\) is a \(1 \times N\) vector of weights (e.g, the
population mean is represented by a weights vector whose elements all
equal \(1 / N\)). Denoting quantities that are part of the sampled
population units with a subscript \emph{s} and quantities that are part
of the unsampled population units with a subscript \emph{u}, let

@@

In this section we describe how we used simulated and real data to
investigate performance between simple random sampling (SRS) and GRTS
sampling as well as performance between design-based (DB) and
model-based (MB) inference. In SRS and GRTS sampling, all population
units had equal inclusion probabilities and were selected without
replacement. The important distinction between SRS and GRTS is that SRS
ignores spatial locations while sampling but GRTS explicitly
incorporates them. Together, the two sampling plans (SRS and GRTS)
combined with the two inference approaches (DB and MB) yielded four
sampling-inference combinations: SRS-DB, SRS-MB, GRTS-DB, and GRTS-MB.
For SRS-DB, the Horvitz-Thompson estimator \eqref{eq:ht} was used to
estimate means and the commonly-used SRS variance formula (Lohr, 2009;
Särndal et al., 2003) was used to estimate variances. This variance
formula is given by \begin{equation}\label{equation:srs_var}
 \frac{f[\sum_{i = 1}^n (z_i - \bar{z})^2]}{n(n - 1)},
\end{equation} where \(z_i\) is the response value of the \(i\)th
sampled population unit, \(\bar{z}\) is the mean of all \(z_i\), \(n\)
is the sample size, \(N\) is the population size, and
\(f = (1 - n / N)\) (\(f\) is often called the finite population
correction factor). For GRTS-DB, the Horvitz-Thompson estimator was used
to estimate means and the local neighborhood variance was used to
estimate variances. For SRS-MB and GRTS-MB, FPBK was used to estimate
means and variances using restricted maximum likelihood. SRS, GRTS
sampling, and design-based inference were implemented using the
\texttt{spsurvey} \textbf{\textsf{R}} package (Dumelle et al., 2022).
FPBK was implemented using the \texttt{sptotal} \textbf{\textsf{R}}
package (Higham et al., 2021a).

@@

\leavevmode\hypertarget{ref-higham2021sptotal}{}%
Higham, M., Ver Hoef, J., Frank, B., Dumelle, M., 2021a. Sptotal:
Predicting totals and weighted sums from spatial data.

\leavevmode\hypertarget{ref-higham2021adjusting}{}%
Higham, M., Ver Hoef, J., Madsen, L., Aderman, A., 2021b. Adjusting a
finite population block kriging estimator for imperfect detection.
Environmetrics 32, e2654.
